Who Gets Cared For and Why? Fertility Shocks, Intra-Household Investment Reallocation, and the Quantity--Quality Trade-Off in the Philippines

### 2025-08-09 docs/paper.Rmd
fix typos and update regression estimate in paper
Corrected a missing period in the abstract and updated the Model 2 coefficient for the first-stage regression (from 0.200 to 0.242) in both the RMarkdown and LaTeX source files. These changes clarify the results and improve the accuracy of the manuscript.

  This paper investigates how increases in fertility, driven by reduced access to contraception, affect parental investment in child health. I use Executive Order No. 003—a 2000 policy in Manila that restricted access to modern contraceptives—as an exogenous fertility shock. Drawing on pooled data from the 1998, 2003, and 2008 Philippine Demographic and Health Surveys, I estimate two-stage least squares models that use policy exposure to instrument for sibship size. Child health inputs fall into two categories: prenatal (proxied by birth weight) and postnatal (measured by the total number of vaccine doses received by age five). The first-stage results indicate that affected mothers had, on average, 0.20 more living children at the time of a birth. In the second stage, each additional sibling is associated with a reduction of 22.97 vaccine doses received by the index child, a decline large enough to prevent completion of the recommended immunization schedule. The estimated effect on birth weight is −8,676 grams, though statistically insignificant Reduced-form estimates are directionally consistent: policy exposure is associated with a 0.91-dose drop in vaccine uptake and a 349.60-gram decrease in birth weight. While instrumental variable estimates are limited by a weak first stage, the magnitude and sign of the coefficients support the hypothesis that larger sibship sizes reduce health investments per child. The results suggest that postnatal investments, particularly those requiring sustained parental effort, are more sensitive to fertility shocks than prenatal outcomes.

@@

Table \ref{tab:first-stage-sibship} presents first-stage regression estimates assessing the relationship between exposure to Manila’s post-2000 fertility-restrictive policy and sibship size. The outcome variables are the number of living children at the time of survey (Model 1) and the total number of children ever born, including deceased (Model 2). In both models, the interaction term Manila × Post-2000 is positive—0.098 (SE = 0.691) in Model 1 and 0.200 (SE = 0.461) in Model 2—indicating that children born to Manila mothers after the policy may have experienced slightly larger sibship sizes, on average. Although these estimates are not statistically significant, their consistent positive direction suggests that the policy did not produce a strong fertility-decreasing effect, and may even have been associated with modest increases in completed or surviving fertility.

>>>>

  This paper investigates how increases in fertility, driven by reduced access to contraception, affect parental investment in child health. I use Executive Order No. 003—a 2000 policy in Manila that restricted access to modern contraceptives—as an exogenous fertility shock. Drawing on pooled data from the 1998, 2003, and 2008 Philippine Demographic and Health Surveys, I estimate two-stage least squares models that use policy exposure to instrument for sibship size. Child health inputs fall into two categories: prenatal (proxied by birth weight) and postnatal (measured by the total number of vaccine doses received by age five). The first-stage results indicate that affected mothers had, on average, 0.20 more living children at the time of a birth. In the second stage, each additional sibling is associated with a reduction of 22.97 vaccine doses received by the index child, a decline large enough to prevent completion of the recommended immunization schedule. The estimated effect on birth weight is −8,676 grams, though statistically insignificant. Reduced-form estimates are directionally consistent: policy exposure is associated with a 0.91-dose drop in vaccine uptake and a 349.60-gram decrease in birth weight. While instrumental variable estimates are limited by a weak first stage, the magnitude and sign of the coefficients support the hypothesis that larger sibship sizes reduce health investments per child. The results suggest that postnatal investments, particularly those requiring sustained parental effort, are more sensitive to fertility shocks than prenatal outcomes.

@@

Table \ref{tab:first-stage-sibship} presents first-stage regression estimates assessing the relationship between exposure to Manila’s post-2000 fertility-restrictive policy and sibship size. The outcome variables are the number of living children at the time of survey (Model 1) and the total number of children ever born, including deceased (Model 2). In both models, the interaction term Manila × Post-2000 is positive—0.098 (SE = 0.691) in Model 1 and 0.242 (SE = 0.461) in Model 2—indicating that children born to Manila mothers after the policy may have experienced slightly larger sibship sizes, on average. Although these estimates are not statistically significant, their consistent positive direction suggests that the policy did not produce a strong fertility-decreasing effect, and may even have been associated with modest increases in completed or surviving fertility.

### 2025-08-09 docs/paper.tex
fix typos and update regression estimate in paper
Corrected a missing period in the abstract and updated the Model 2 coefficient for the first-stage regression (from 0.200 to 0.242) in both the RMarkdown and LaTeX source files. These changes clarify the results and improve the accuracy of the manuscript.

\begin{abstract}
This paper investigates how increases in fertility, driven by reduced
access to contraception, affect parental investment in child health. I
use Executive Order No.~003---a 2000 policy in Manila that restricted
access to modern contraceptives---as an exogenous fertility shock.
Drawing on pooled data from the 1998, 2003, and 2008 Philippine
Demographic and Health Surveys, I estimate two-stage least squares
models that use policy exposure to instrument for sibship size. Child
health inputs fall into two categories: prenatal (proxied by birth
weight) and postnatal (measured by the total number of vaccine doses
received by age five). The first-stage results indicate that affected
mothers had, on average, 0.20 more living children at the time of a
birth. In the second stage, each additional sibling is associated with a
reduction of 22.97 vaccine doses received by the index child, a decline
large enough to prevent completion of the recommended immunization
schedule. The estimated effect on birth weight is −8,676 grams, though
statistically insignificant Reduced-form estimates are directionally
consistent: policy exposure is associated with a 0.91-dose drop in
vaccine uptake and a 349.60-gram decrease in birth weight. While
instrumental variable estimates are limited by a weak first stage, the
magnitude and sign of the coefficients support the hypothesis that
larger sibship sizes reduce health investments per child. The results
suggest that postnatal investments, particularly those requiring
sustained parental effort, are more sensitive to fertility shocks than
prenatal outcomes.
\end{abstract}

@@

Table \ref{tab:first-stage-sibship} presents first-stage regression
estimates assessing the relationship between exposure to Manila's
post-2000 fertility-restrictive policy and sibship size. The outcome
variables are the number of living children at the time of survey (Model
1) and the total number of children ever born, including deceased (Model
2). In both models, the interaction term Manila × Post-2000 is
positive---0.098 (SE = 0.691) in Model 1 and 0.200 (SE = 0.461) in Model
2---indicating that children born to Manila mothers after the policy may
have experienced slightly larger sibship sizes, on average. Although
these estimates are not statistically significant, their consistent
positive direction suggests that the policy did not produce a strong
fertility-decreasing effect, and may even have been associated with
modest increases in completed or surviving fertility.

>>>>

\begin{abstract}
This paper investigates how increases in fertility, driven by reduced
access to contraception, affect parental investment in child health. I
use Executive Order No.~003---a 2000 policy in Manila that restricted
access to modern contraceptives---as an exogenous fertility shock.
Drawing on pooled data from the 1998, 2003, and 2008 Philippine
Demographic and Health Surveys, I estimate two-stage least squares
models that use policy exposure to instrument for sibship size. Child
health inputs fall into two categories: prenatal (proxied by birth
weight) and postnatal (measured by the total number of vaccine doses
received by age five). The first-stage results indicate that affected
mothers had, on average, 0.20 more living children at the time of a
birth. In the second stage, each additional sibling is associated with a
reduction of 22.97 vaccine doses received by the index child, a decline
large enough to prevent completion of the recommended immunization
schedule. The estimated effect on birth weight is −8,676 grams, though
statistically insignificant. Reduced-form estimates are directionally
consistent: policy exposure is associated with a 0.91-dose drop in
vaccine uptake and a 349.60-gram decrease in birth weight. While
instrumental variable estimates are limited by a weak first stage, the
magnitude and sign of the coefficients support the hypothesis that
larger sibship sizes reduce health investments per child. The results
suggest that postnatal investments, particularly those requiring
sustained parental effort, are more sensitive to fertility shocks than
prenatal outcomes.
\end{abstract}

@@

Table \ref{tab:first-stage-sibship} presents first-stage regression
estimates assessing the relationship between exposure to Manila's
post-2000 fertility-restrictive policy and sibship size. The outcome
variables are the number of living children at the time of survey (Model
1) and the total number of children ever born, including deceased (Model
2). In both models, the interaction term Manila × Post-2000 is
positive---0.098 (SE = 0.691) in Model 1 and 0.242 (SE = 0.461) in Model
2---indicating that children born to Manila mothers after the policy may
have experienced slightly larger sibship sizes, on average. Although
these estimates are not statistically significant, their consistent
positive direction suggests that the policy did not produce a strong
fertility-decreasing effect, and may even have been associated with
modest increases in completed or surviving fertility.
